SVG: The Paper

### 2022-05-04 Analysis/index_prpic.Rmd
Starting work on Ppric

```{r, echo = TRUE}
# read in data
d <- read_csv("data/clarke_2020_qjep.csv", show_col_types = FALSE) %>%
  mutate(condition = as_factor(condition),
         condition = fct_recode(condition, feature = "1", conjunction = "2"),
         targ_type = as_factor(targ_type))
```

@@

```{r echo = TRUE}
# item weights pre-computed - see scripts/extract_item_weights_using_model.R
a <- readRDS("scratch/qjep_model_weights.rda") %>%
  filter(found != 1) # remove initial selections

@@

Let's plot some examples oh highly predictable (top 5%), typical (middle 10%) and unpredictable (bottom 5%) trials.

```{r, echo = TRUE}
qs <- quantile(a_trl$prop_max, c(0.05, 0.40, 0.60, 0.95))

trls_hard <- filter(a_trl, prop_max < qs[1]) %>% sample_n(3)
trls_mid  <- filter(a_trl, prop_max > qs[2], prop_max < qs[3]) %>% sample_n(3)
trls_easy <- filter(a_trl, prop_max > qs[4]) %>% sample_n(3)
```

We can see that for these highly predictable trials, we make very few mistakes, and those that we do make appear to be related to some sort of shortest-path optimization on the part of the participant.

```{r fig.cap = "A random selection of *good* trials. When participants diverge from the model's prediction, it appears to be due to some form of path-length optimisation."}

source("functions/plot_trial.R")
plts1 <- map(1:3, plot_trial, trls_easy, a, d)

plts1[[1]] + plts1[[2]] +
  plot_layout(nrow = 1, widths = c(1,1)) &
   theme(plot.margin = unit(c(.2,.2,.2,.2), "cm")) &
   theme(plot.background = element_rect(color  = "#fffff8", size = 2,linetype = 'dotted', fill ="#fffff8")) &
   theme(text=element_text(size=14,  family="serif"))
```

Here are some typical trials. Possible we have the path-length minimization issue again.

```{r fig.cap = "A random selection of *typical* trials. Along with path-minimising behaviour, we can also see some disagreement around switching item type."}
plts2 <- map(1:3, plot_trial, trls_mid, a, d)

plts2[[1]] + plts2[[2]] +
  plot_layout(nrow = 1, widths = c(1,1)) &
   theme(plot.margin = unit(c(.2,.2,.2,.2), "cm")) &
   theme(plot.background = element_rect(color  = "#fffff8", size = 2,linetype = 'dotted', fill ="#fffff8")) &
   theme(text=element_text(size=14,  family="serif"))
```

Now let's look at less predictable (bottom 5%-tile) trials. Even here, the model does a relatively good job.

```{r fig.cap = "A random selection of *hard* trials. It is harder to make out much of a pattern here, but we can see that the human participants jump around more."}
plts3 <- map(1:3, plot_trial, trls_hard, a, d)

plts3[[1]] + plts3[[2]] +
  plot_layout(nrow = 1, widths = c(1,1)) &
   theme(plot.margin = unit(c(.2,.2,.2,.2), "cm")) &
   theme(plot.background = element_rect(color  = "#fffff8", size = 2,linetype = 'dotted', fill ="#fffff8")) &
   theme(text=element_text(size=14,  family="serif"))
```

```{r}
ggsave("../Figures/qjep_ex_paths.pdf",
      ( plts1[[1]] + plts1[[2]]) /( plts3[[1]] + plts3[[3]]), 
       width = 8, height = 8)

rm(trls_easy, trls_mid, trls_hard, plts, qs, plts1, plts2, plts3)
```

# Path Length Analysis

It looks like some of the target selections that our model fails to capture are when participants select a further away target in order to minimise overall path length. Here we will compare our model's predicted path length against the empirical paths.

- We will provide the model/simulation with the initial target selection.

```{r, fig.height=3.5, fig.cap = "(*left*) Histogram comparing the empirical path lengths to simulations from our fitted model. Difference indicates the human path length - the simulated path length. (*right*) Histogram comparing the median path lengths per participant.", message=FALSE}
path_lengths <- readRDS("scratch/path_lengths.rda") %>%
  rename(observer = "obs",
         condition = "cond",
         human = "human_path_length",
         model = "model_path_length") %>% 
  mutate(difference  = human - model) %>%
  pivot_longer(c(human, model, difference), names_to = "type", values_to = "path_length")

ggplot(path_lengths, aes(path_length, fill = type)) + 
  geom_histogram(bins = 25, alpha = 0.5, position = "identity") +
  geom_vline(xintercept = 0, linetype = 2) + 
  theme(legend.position = "bottom")  +
  scale_x_continuous("path length")  +
  scale_x_continuous("median path length") +
  scale_fill_manual(values = col3) -> plt1

path_lengths %>% group_by(observer, type) %>%
  summarise(path_length = median(path_length), .groups = "drop") %>%
  ggplot(aes(path_length, fill = type)) + 
  geom_histogram(bins = 25, alpha = 0.5, position = "identity") +
  geom_vline(xintercept = 0, linetype = 2) + 
  theme(legend.position = "bottom") +
  scale_x_continuous("median path length") +
  scale_fill_manual(values = col3) -> plt2

plt1 + plt2 + plot_layout(guides = "collect")  &
  theme(legend.position = 'bottom', legend.title = element_blank())
```

We can see that overall, the model is not making substantially longer paths through the items than our human participants. However, there is some variability in that some participants routinely select paths shorter than our model would, and others do not.

Perhaps this is related to predictability (i.e., how well the model can predict a participant's trial level behaviour)?

```{r fig.height=3, fig.cap = "(*left*) A histogram showing the proportion of trials each participant managed a shorter path through the targets than the model's prediction. (*right*) Scatter plot showing the relationship between predictability (accuracy) and whether the model over or under estimates path length. Xs mark the outlier participants discussed above."}
path_lengths %>% filter(type == "difference") %>%
  mutate(shorter = path_length < 0) %>%
  group_by(observer, condition) %>%
  summarise(prop_shorter = mean(shorter), .groups = "drop") -> path_lengths_obs

path_lengths_obs %>%
  ggplot(aes(prop_shorter, fill = condition)) + 
  geom_histogram(alpha = 0.5, breaks = seq(0, 1, 0.2), position = "identity") +
  scale_x_continuous("prop. shorter than model", breaks = seq(0, 1, 0.2)) -> plt1

path_lengths <- full_join(path_lengths_obs, fit,
                          by = c("observer", "condition"))

ggplot(path_lengths, aes(accuracy, prop_shorter, colour = condition)) + 
  geom_point(aes(shape = is_outlier)) +
  geom_smooth(data = filter(path_lengths, is_outlier == 0), 
              method = "lm", formula = y ~ x, se = FALSE) + 
  scale_shape_manual(values = c(19, 4), guide = "none") + 
  scale_y_continuous("prop. shorter than model", breaks = seq(0, 1, 0.2)) +
  coord_cartesian(ylim = c(0, 1)) -> plt2

plt1 + plt2 + plot_layout(guides = "collect")  &
  theme(legend.position = 'bottom')
```

```{r}
rm(path_lengths, plt1, plt2)
```

# Initial Selection

Previous work (cite Clarke and Tatler etc) has demonstrated that the initial fixation in scene viewing can be well described with truncated Gaussian distribution. Here we explore using a similar idea to account for starting locations.

```{r echo = TRUE}
# Reimport data again (as we removed the initial target selection earlier)
d <- read_csv("data/clarke_2020_qjep.csv", show_col_types = FALSE) %>%
  filter(found == 1) %>%
  mutate(condition = as_factor(condition))
```

```{r fig.height = 3.5, fig.cap="(*left*) Density plot for initial target selections, over all participants and all trials. (*right*) Median $x$ and $y$ coordinates for the initial target selection for each participant."}
my_theme <- theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_rect(color  = "#808080", size = 2,linetype = 'dotted', fill ="#808080"),
        panel.border= element_rect(colour = "black", fill=NA, size=5),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank())

ggplot(d, aes(x, y)) + geom_hex(aes(colour = ..count..), bins = 12) +
  scale_fill_viridis_c() + scale_color_viridis_c() +
  my_theme -> plt_all

d %>% group_by(observer, condition) %>% 
  summarise(x = median(x), y = median(y),
            .groups = "drop") %>%
  ggplot(aes(x, y)) + geom_hex(bins = 12) +
  scale_fill_viridis_c()+ scale_color_viridis_c() +
  my_theme -> plt_med

plt_all + plt_med

rm(plt_all, plt_med)
```

We can model this as multi-level Gaussian? Yes, although it seems like people are in one of two clusters.

```{r fig.cap = "The mean posterior plot for the truncated normal distriubtions for (*left*) $x$ and (*right*) $y$. We can see that a two component mixture model will capture most of the variance in intial target selection."}

m <- readRDS("scratch/init_sel_model.rds")
z <- seq(0, 1, 0.01)

gather_draws(m, u[dim, observer]) %>% 
  ungroup() %>%
  select(-.variable) %>%
  rename(obs_b = ".value") %>%
  full_join(gather_draws(m, b[dim]),
             by = c("dim", ".chain", ".iteration", ".draw")) %>%
  mutate(obs_b = .value + obs_b,
         .variable = if_else(dim %in% c(1,2), "mu", "sd"),
         dim = if_else(dim %in% c(1,3), "x", "y")) %>%
  select(observer, dim, .iteration, obs_b, param = ".variable") %>%
  mutate(obs_b = if_else(param == "sd", exp(obs_b), obs_b)) -> a

a %>% group_by(observer, param, dim) %>%
  summarise(value = mean(obs_b) , .groups = "drop") %>%
  pivot_wider(names_from = param, values_from = value) %>%
  rename(mean = "mu") -> am

create_post_plot <- function(dm) {

  pmap(filter(am,  dim == dm) %>% select(mean, sd), dtruncnorm, a = 0, b= 1, x = z) %>%
    set_names(rownames(filter(am,  dim == dm))) %>%
    bind_rows() %>%
    mutate(z = z) %>%
    pivot_longer(-z, names_to = "observer", values_to = "v") %>%
    ggplot(aes(x = z, y = v, group = observer)) + geom_path(alpha = 0.25) +
    scale_x_continuous(dm) + scale_y_continuous("likelihood") -> plt

  return(plt)
}

create_post_plot("x") + create_post_plot("y")
```

- Do we want to simplify this model to something in which people have some mix of a (fixed) centre and corner bias?

Do participants who differ in terms of their initial target selection differ in other ways too?

```{r echo = TRUE, fig.cap = "Whether participants have a central or corner bias for the intiial target selection appears to be largely independent of their other model parameters."}
full_join(fit, filter(am, dim == "x"), by = "observer") %>%
  pivot_longer(c(pA, pS, bP, bM, accuracy), 
               names_to = "statistic", values_to = "value") %>%
  mutate(intial_sel = if_else(mean < 0, "corner", "central")) %>%
  ggplot(aes(statistic, value, fill = intial_sel)) +
  geom_boxplot(alpha = 0.5) + 
  facet_wrap(~statistic, scales = "free", nrow = 1) +
  theme(legend.position = "bottom") + 
   scale_fill_manual("intial target selection",
                     values = c(col3[2], col3[3]))
```

# Item Density

>>>>

```{r, echo = TRUE}
# read in data
d <- read_csv("data/squirrel_spatial.csv", show_col_types = FALSE) %>%
  mutate(condition = as_factor(condition),
         condition = fct_recode(condition, feature = "1", conjunction = "2"),
         targ_type = as_factor(targ_type))
```

@@

```{r echo = TRUE}
# item weights pre-computed - see scripts/extract_item_weights_using_model.R
a <- readRDS("scratch/prpic_model_weights.rda") %>%
  filter(found != 1) # remove initial selections

@@
